Inferring ecological interactions

### 2017-03-02 ms/ms_interactions.tex
Reorganized the co-occurrence models section

When modelling ecological networks, there are different types of methods that
have been proposed in the litterature. In this section, we presented the
different techniques and group them based on their properties.

\subsubsection*{Co-occurrences models}
\label{regression-based-models}
%%% Regression-based models on co-occurrences subsubsection will probably have to be moved somewhere else to follow the structure of the Introduction where traits and phylogeny related methods are discussed firsts and co-occurrence methods are discussed later.

The simplest model of that can be designed to characterize the co-occurrence
patterns between species $i$ and $j$ in matrix $Y$ can be defined as follow

\begin{equation}
  \label{eq:coOccSimple}
  Y_{i,j} = \rho_{i,j}
\end{equation}

where the probability of interaction among species $i$ and $j$ is a fixed value
($\rho_{i,j}$). Using empirical data, we might try to estimate this value by
dividing the number of observations by the total number of observations for any
pair of species. While this is straightforward, it is insufficient because it
contains no estimate of uncertainty. The estimate is given without any
consideration for the variance due to sampling, environmental, spatial or
temporal stochasticity. This model is overly simple. By assuming that
observations are randomly drawn from a statistical distribution with parameter
$\rho_{i,j}$, we obtain a simple model that account for sampling stochasticity
in the data.

\begin{equation}
  \label{eq:coOccBern}
  Y_{i,j} \sim \text{Bernoulli}(\rho_{i,j})
\end{equation}

In equation \ref{eq:coOccBern}, we used a Bernoulli distribution, which assumes
that $Y_{i,j}$ is either 0 or 1. If $Y_{i,j}$ includes other types of data (e.g.
counts or weights) other distributions may be used instead. From equation
\ref{eq:coOccBern}, we can now estimate the probability of interaction while
accounting sampling stochasticity. Although this model is more informative than
the one presented in equation \ref{eq:coOccSimple}, it is still unsatisfying
because it depends wholly on species identity and uses no ecological theory or
inference; in short it is not mechanistic. Therefore, it is difficult to ascribe
meaning to the species association that could be highlighted by this model, or
understand how they might change in the future. In addition, from this model, we
cannot make a prediction about potential new species.

To make our model more ecologically relevant, we might add covariates $X_{i,j}$
that influence the relationship among a pair of species

\begin{equation}
  \label{eq:coOccCov}
  Y_{i,j} \sim \text{Bernoulli}(\rho_{i,j})
  \text{logit}(\rho_{i,j}) = \alpha_{i,j}+\beta_{i,j}X_{i,j}
\end{equation}

where $\beta_{i,j}$ is a regression coefficient and $\alpha_{i,j}$ is the model
intercept. In equation \ref{eq:coOccCov}, a link between species $i$ and $j$ is
a Bernoulli trial with a probability of success $\rho_{i,j}$. The probability of
success will vary based on some ecological principle ($x_{i,j}$) such as the
similarity in functional traits among species $i$ and $j$:

%%% GB: I am not sure if this equation is right... Need to check carefully.
\begin{equation}
  \label{eq:coOccTrait}
  x_{i,j}= f(\text{Trait}_i,\text{Trait}_j)
\end{equation}

%%% ICI %%%

A variety of trait-matching functions may be biologically plausible. For
example, the absolute value of the difference between traits (e.g. \(|trait_i -
trait_j|\)) or a binary difference (e.g. \(trait_i > trait_j\)), could be used
to predict which species may interact. Alternatively, the covariate may
represent the phylogenetic relatedness of species, or their relative response to
a given environmental variable.In some cases, we cannot provide a simple
trait-matching function (eg V), but must use a statistical tool to infer the
best possible combination of traits among co-occurring species that predicts
interactions. Machine learning tools provide an ideal compliment to traditional
linear models when we cannot anticipate the combinations of traits that can
influence species interaction rates. Unsupervised learning tools, such as random
forests, can give us the associations among species given a set of complex
traits. Regardless of the exact form of regression, this framework satisfies two
important goals: it allows for estimates of uncertainty in species interactions,
and it provides a mechanism shaping the probability of an interaction.

So far, we have only captured the uncertainty in the process model, the inherent
temporal or spatial stochasticity in our ecological process of interest (Hooten
and Hobbs). We may need to also account for the uncertainty in the observation
model, the ability to detect interactions given that they occur. The separation
of the observation and process model translates our ecological mechanism into
specific predictions about empirical data. For binary networks, one
straightforward way to account for the observation of interactions is to model
the detectability of network interactions.

eq iv.

\[ Y_{i,j,k} \sim Bernoulli(\phi_{i,j})\] \[ \phi_{i,j} = \omega_{i,j} *
z_{i,j}\] \[z_{i,j} \sim Bernoulii(\rho_{i,j})\] \[ logit(\rho_{i,j}) =
\alpha_{i,j} + \beta_{i,j} * x \]

Where the observation of a link between species i and species j at time k is a
Bernoulli trial with a probability \(\phi_{i,j}\), which is the outcome of the
detectability of an interaction (\(\omega_{i,j}\)) and the latent state
(\(z_{i,j}\)). This latent state is the true, but unobserved, existence of a
link, as predicted by our ecological mechanism of interest (see eq IV). Our
ability to detect this latent state depends on the detectability of an
interaction (\(\omega_{i,j}\)). The benefit of this approach is that
differentiates the probability of detection from the probability of occurrence.
Importantly, because we have explicitly defined some temporal window (k),
interactions with different levels of sampling effort can be directly compared.

@@

\subsubsection*{Non-parametric methods}\label{non-parametric-methods}

The above methods are model based, that is, they largely use parametric
distributions to estimate the likelihood of the data given estimated model
parameters. An alternative approach is to use multivariate statistics to create
association matrices. One such method is called fourth-corner analysis, which is
a type of three table ordination (sometimes called `RLQ' analysis) (Doledec
1996). In bipartite networks, fourth-corner analysis relates the matrix of
species interactions to a matrix of traits from one level (e.g.~plants) to
another level (e.g.~pollinators) via the matrix of species innteractions (Dray \&
Legendre, 2008). This produces a correlation matrix, which is then compared to a
null expectation through matrix randomization. Randomization is a familiar tool
in network ecology, in which permuatation rules are used to compare the observed
test statistic (e.g the correlation among species occurrences) to a null
distribution based on some matrix constraint. Common constraints include the
mantainance of row sums, column sums, or other all marginal values, such that
only the identity of partners are switched. After each permutation, the test
statistic is recalulcating, yielding a null distribution. By comparing whether
the observed test statistic falls within the upper or lower 5th quantile of the
null distribution, studies distinguish whether their results could have been
generated at random given their sampling effort.

>>>>

When modelling ecological networks, there are different types of methods that
have been proposed in the literature. In this section, we presented the
different techniques and group them based on their properties.

\subsubsection*{Co-occurrences models}
\label{Co-occurrences models}

% Regression-based models on co-occurrences subsubsection will probably have
% to be moved somewhere else to follow the structure of the Introduction where
% traits and phylogeny related methods are discussed firsts and co-occurrence
% methods are discussed later.

Co-occurence models are designed to characterize species co-occurrence directly.
In these models, the co-occurrences $Z_{ij}$ we try to characterize describes
the presence (1) or absence (0) of a trophic link between species $i$ and $j$.
In the context of co-occurrence models,  a way to formalize the relationship
between co-occurrences and traits is as follow

\begin{equation}
  \label{eq:coOccBern}
  Z_{ij} \sim \text{Bernoulli}(\rho_{ij})
\end{equation}

where $\rho_{ij}$ is the probability of interaction among species $i$ and $j$.
In equation \ref{eq:coOccBern}, we used a Bernoulli distribution because we
assumed that $Z_{ij}$ is binary. However, if $Z_{ij}$ is characterized by other
types of data (e.g. counts or weights) other distributions may be used instead.
From equation \ref{eq:coOccBern}, we can estimate the probability of interaction
while accounting for sampling stochasticity. However, because this model depends
solely on species identity it may lead to misinterpretation of the mechanisms
that characterize the co-occurrences.

A way to solve this problem is to add a layer of complexity to the model in
equation \ref{eq:coOccBern} by including traits. To do this, the probability of
interaction between species $i$ and $j$ defined by $\rho_{ij}$ can depend on the
similarity in functional traits for species $i$ and $j$

\begin{equation}
  \label{eq:coOccTraits}
  \text{logit}(\rho_{ij}) = \beta_{ij}X_{ij}
\end{equation}

where $\beta_{ij}$ is a regression coefficient and $X_{ij}$ is a covariates that
influence the relationship among a pair of species. For example, $X_{ij}$ could
be a trait-matching function between species $i$ and $j$. Note that a model
intercept is accounted for in equation \ref{eq:coOccTraits} by assuming that $X$
includes a column of 1s. In equation \ref{eq:coOccTraits}, we used a logit link
function to ensure that $\rho$ ranges between 0 and 1, however, it is also
possible to use other link functions such as a probit or a complementary log-log
link function.

A variety of trait-matching functions may be biologically plausible. For
example, the absolute value of the difference between traits or a binary
difference, can be used to predict which species may interact. Alternatively,
the covariate may also represent phylogenetic relatedness, or their relative
response to a given environmental variable.

In the litterature, \#\#\#REF(Ben?)\#\#\# also proposed to account for
observation uncertainty; to account for the ability to detect interactions given
that they really occurred. In this formulation, the idea is to try to seperate
the error resulting from the observation and the processes defining the
co-occurrences. As such, we can define a model that account for observation uncertainty and ecological processes through the following equations

\begin{align}
  Z_{ijk} \sim \text{Bernoulli}(\phi_{ij})\\
  \phi_{ij} = \omega_{ij} \times \Omega_{ij}\\
  \Omega_{ij} \sim \text{Bernoulli}(\rho_{i,j})\\
  logit(\rho_{ij}) = \beta_{ij} X_{ij}
\end{align}

where the observation at time $k$ of a co-occurrence between species $i$ and $j$
is a Bernoulli trial with a probability $\phi_{ij}$, which is the outcome of the
detectability of an interaction ($\omega_{ij}$) and the latent state
($\Omega_{ij}$). The latent state $\Omega_{ij}$ is the true, but unobserved,
existence of a co-occurrence, as predicted by the ecological mechanism of
interest. The ability to assess this latent state depends on the detectability
of an interaction ($\omega_{i,j}$).

\subsubsection*{fourth-corner approaches}
\label{fourth-corner approaches}

A way to assess

An alternative approach is to use multivariate statistics to create association
matrices. One such method is called fourth-corner analysis, which is a type of
three table ordination (sometimes called `RLQ' analysis) (Doledec 1996). In
bipartite networks, fourth-corner analysis relates the matrix of species
interactions to a matrix of traits from one level (e.g.~plants) to another level
(e.g.~pollinators) via the matrix of species innteractions (Dray \& Legendre,
2008). This produces a correlation matrix, which is then compared to a null
expectation through matrix randomization. Randomization is a familiar tool in
network ecology, in which permuatation rules are used to compare the observed
test statistic (e.g the correlation among species occurrences) to a null
distribution based on some matrix constraint. Common constraints include the
mantainance of row sums, column sums, or other all marginal values, such that
only the identity of partners are switched. After each permutation, the test
statistic is recalulcating, yielding a null distribution. By comparing whether
the observed test statistic falls within the upper or lower 5th quantile of the
null distribution, studies distinguish whether their results could have been
generated at random given their sampling effort.

%%% I think these bits of text would fit better in the discussion
%% The text in 1) were all from the co-occurrence model section.
% 1.1)
In some cases, we cannot provide a simple trait-matching function but must use a
statistical tool to infer the best possible combination of traits among
co-occurring species that predicts interactions. Machine learning tools provide
an ideal compliment to traditional linear models when we cannot anticipate the
combinations of traits that can influence species interaction rates.
Unsupervised learning tools, such as random forests, can give us the
associations among species given a set of complex traits. Regardless of the
exact form of regression, this framework satisfies two important goals: it
allows for estimates of uncertainty in species interactions, and it provides a
mechanism shaping the probability of an interaction.

% 1.2)
The benefit of this approach is that differentiates the probability of detection
from the probability of occurrence. Importantly, because we have explicitly
defined some temporal window (k), interactions with different levels of sampling
effort can be directly compared.

@@
